MSstatsPTM statistical relative quantification of post-translational modifications in bottom-up proteomic experiments

### 2022-04-06 main/MSstatsPTM_Main.tex
updated notation

Figure \ref{fig:data-structure}(a) schematically illustrates this data structure for one PTM and corresponding unmodified protein. The number of quantified features vary across replicate LC-MS/MS runs of the same sample, and across conditions. There may be feature outliers which could be caused by random variation or measurement error. Additionally, some spectral features can be missing. The features in each mass spectrometry run are generally summarized and represent a single observation used to estimate the relative abundance of the PTM in condition $i$, $u_i^{\ast}$.

The population parameter we are interested in estimating is the difference between the PTM in condition $1$ and condition $2$, denoted by $u_1^{\ast}$ and $u_2^{\ast}$ respectively. To calculate the quantity of interest, statistical analysis needs to summarize values in the table in Figure \ref{fig:data-structure}(a) using appropriate models. For condition $i$, the models provide estimates for the mean abundance, $\hat{u}_i^{\ast}$, and standard error, $\hat{SE}(\hat{u}_i^{\ast})$. Once summarized, we can formulate the Null and Alternative Hypotheses as seen in Equation \ref{eq:conv_null_hyp}.

@@

However, these population parameters are inherently convoluted with changes in the unmodified protein. To account for this, we must estimate the relative abundance of the unmodified protein in conditions $1$ and $2$, denoted by $\mu_1$ and $\mu_2$. We estimate these quantities in the same way as for the PTM, using their correspond features and observations and resulting in an estimated mean abundance, $\hat{u}_i$, and standard error, $\hat{SE}(\hat{u}_i)$. These estimates are used to adjust the quantities estimated for the PTM, resulting in the new Null and Alternative Hypotheses seen in Equation \ref{eq:null_hyp}.

@@

For adjustment with respect to unmodified peptides, the estimate of PTM abundance is divided by the protein abundance estimate, which is equivalent to take the difference of their log abundances. For condition $i$ and biological replicate $j$, the quantity is denoted by $\hat{u}^{\ast}_{ij}$ and is given by Equation \ref{eq:adj_estimation}.

@@

$t$-test then takes as input the adjusted PTM abundance of each condition and biological replicate, $\hat{u}^{\ast}_{11}$, $\hat{u}^{\ast}_{12}$, $\hat{u}^{\ast}_{21}$, and $\hat{u}^{\ast}_{22}$, to estimate the quantity of interest $\hat{\Delta}$ as well as the standard error, $\hat{SE}(\hat{\Delta})$. $\hat{\Delta}$ is the estimate of the difference for the Null and Alternative Hypotheses show in Equation \ref{eq:null_hyp}. The statistical significance is determined by comparing the test statistic against the $t$ distribution, with degrees of freedom $df=2J-2$ in balanced designs.

\medskip \noindent {\bf Limma} Modifications of $t$-test such as moderated $t$-test with Limma were also proposed~\cite{Ritchie_15a}\cite{Zhu}. The Null Hypothesis we are interested in answering is the same as in Equation \ref{eq:null_hyp}. To estimate $\hat{\Delta}$ in Figure \ref{fig:data-structure}, feature level summarization and unmodified protein adjustment are performed in the same way as $t$-test, shown in Equations \ref{eq:feature_estimation} and \ref{eq:adj_estimation}. Limma then fits a linear model using the same adjusted summarized features as in $t$-test, $\hat{u}^{\ast}_{11}$, $\hat{u}^{\ast}_{12}$, $\hat{u}^{\ast}_{21}$ and $\hat{u}^{\ast}_{22}$, to estimate $\hat{\Delta}$. To estimate $\hat{SE}(\hat{\Delta})$, Limma leverages Empirical Bayes moderation to moderate the model variance of the individual PTM using the shared global variance of all PTMs modeled in the experiment. For a single PTM $m$ with an estimated fold change $\hat{\Delta}_m$, the estimated variance is calculated as $s^{adj}_{m} = P(s_{m}| s_\Theta)$ where $s_{m}$ is the variance resulting from the linear model of PTM $m$ and $s_\Theta$ is the global variance across all PTMs. The statistical significance is determined using $\hat{\Delta}_m$ and $\hat{SE}(\hat{\Delta})$ calculated using the adjusted variance $s^{adj}_{m}$.

@@

The proposed approach leverages the large amount of work done to create $MSstats$~\cite{Choi:2014}. $MSstats$ was designed for protein significance analysis and statistical relative quantification of proteins and peptides in global, targeted and data-independent proteomics. In terms of Figure \ref{fig:data-structure}, $MSstats$ was designed to estimate the difference in the unmodified protein in condition $1$ and condition $2$, denoted by $u_1$ and $u_2$ respectively. In other words it allows us to answer the new Null Hypothesis seen in Equation \ref{eq:msstats_null_hyp}.

@@

We propose a statistical method for detecting changes in PTMs that allows us to answer the Null Hypothesis stated in Equation \ref{eq:null_hyp}; there is no difference in log-abundance of the PTM site between conditions, adjusted with respect to unmodified protein abundance. It leverages the methods used in $MSstats$ to answer this hypothesis, while being applicable to experiments with complex designs. Relating to the example in Figure \ref{fig:data-structure}, the expected value of the log-abundance of the PTM and unmodified protein in Condition $1$ are denoted by $\hat{\mu}_{1}^{\ast}$ and $\hat{\mu}_{1}$ respectively. The feature level summarization for Condition $1$ is denoted as $\hat{y}_{1+}^{\ast}$ for the modified peptide and $\hat{y}_{1+}$ for the unmodified peptide.

To calculate $\hat{y}_{1+}^{\ast}$ and $\hat{y}_{1+}$, feature and run level summarization is carried out as in the sub-plot model of MSstats using Tukey's median polish. The summarized values for the PTM and unmodified protein are then used to fit two models based on the family of linear mixed-effects models. Finally, the adjusted difference, $\hat{\Delta}$, is given by the difference in log-abundance of the PTM site, subtracted by the difference in log-abundance of the underlying protein, which is equivalent to the log of the ratio of PTM abundance difference to protein abundance difference.

\begin{equation}
\hat{\Delta} = \left[ \frac{1}{J} \left( \hat{y}_{1+}^{\ast} - \hat{y}_{2+}^{\ast} \right) \right] - \left[ \frac{1}{J} \left( \hat{y}_{1+} - \hat{y}_{2+} \right) \right]
\label{eq:prop_fc}
\end{equation}

@@

\begin{equation}
SE(\hat{\Delta}) = \left[ \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]^{1/2}
\label{eq:prop_se}
\end{equation}

@@

After using the converters, the next step is peptide/protein summarization and missing value imputation. For the modified run, the package summarizes features which include the same modification together. Features with multiple modification sites are not included with single site features and are summarized separately. For the global profiling run all unmodified features from the same protein are summarized together, up to the protein level. Additionally, the summarization includes global median normalization and normalizes between MS runs. The package can optionally impute missing values using an AFT model.

>>>>

Figure \ref{fig:data-structure}(a) schematically illustrates this data structure for one PTM and corresponding unmodified protein. The number of quantified features vary across replicate LC-MS/MS runs of the same sample, and across conditions. There may be feature outliers which could be caused by random variation or measurement error. Additionally, some spectral features can be missing. The features in each mass spectrometry run are generally summarized and represent a single observation used to estimate the relative abundance of the PTM in condition $i$, $\mu_i^{\ast}$.

The population parameter we are interested in estimating is the difference between the PTM in condition $1$ and condition $2$, denoted by $\mu_1^{\ast}$ and $\mu_2^{\ast}$ respectively. To calculate the quantity of interest, statistical analysis needs to summarize values in the table in Figure \ref{fig:data-structure}(a) using appropriate models. For condition $i$, the models provide estimates for the mean abundance, $\hat{\mu}_i^{\ast}$, and standard error, $\widehat{SE(\mu_i^{\ast})}$. Once summarized, we can formulate the Null and Alternative Hypotheses as seen in Equation \ref{eq:conv_null_hyp}.

@@

However, these population parameters are inherently convoluted with changes in the unmodified protein. To account for this, we must estimate the relative abundance of the unmodified protein in conditions $1$ and $2$, denoted by $\mu_1$ and $\mu_2$. We estimate these quantities in the same way as for the PTM, using their correspond features and observations and resulting in an estimated mean abundance, $\hat{\mu}_i$, and standard error, $\widehat{SE(\hat{\mu}_i)}$. These estimates are used to adjust the quantities estimated for the PTM, resulting in the new Null and Alternative Hypotheses seen in Equation \ref{eq:null_hyp}.

@@

For adjustment with respect to unmodified peptides, the estimate of PTM abundance is divided by the protein abundance estimate, which is equivalent to take the difference of their log abundances. For condition $i$ and biological replicate $j$, the estimated quantity is denoted by $\hat{u}^{\ast}_{ij}$ and is given by Equation \ref{eq:adj_estimation}.

@@

$t$-test then takes as input the adjusted PTM abundance of each condition and biological replicate, $\hat{u}^{\ast}_{11}$, $\hat{u}^{\ast}_{12}$, $\hat{u}^{\ast}_{21}$, and $\hat{u}^{\ast}_{22}$, to estimate the quantity of interest $\hat{\Delta}$ as well as the standard error, $\widehat{SE(\hat{\Delta})}$. $\hat{\Delta}$ is the estimate of the difference for the Null and Alternative Hypotheses show in Equation \ref{eq:null_hyp}. The statistical significance is determined by comparing the test statistic against the $t$ distribution, with degrees of freedom $df=2J-2$ in balanced designs.

\medskip \noindent {\bf Limma} Modifications of $t$-test such as moderated $t$-test with Limma were also proposed~\cite{Ritchie_15a}\cite{Zhu}. The Null Hypothesis we are interested in answering is the same as in Equation \ref{eq:null_hyp}. To estimate $\hat{\Delta}$ in Figure \ref{fig:data-structure}, feature level summarization and unmodified protein adjustment are performed in the same way as $t$-test, shown in Equations \ref{eq:feature_estimation} and \ref{eq:adj_estimation}. Limma then fits a linear model using the same adjusted summarized features as in $t$-test, $\hat{u}^{\ast}_{11}$, $\hat{u}^{\ast}_{12}$, $\hat{u}^{\ast}_{21}$ and $\hat{u}^{\ast}_{22}$, to estimate $\hat{\Delta}$. To estimate $\widehat{SE(\hat{\Delta})}$, Limma leverages Empirical Bayes moderation to moderate the model variance of the individual PTM using the shared global variance of all PTMs modeled in the experiment. For a single PTM $m$ with an estimated fold change $\hat{\Delta}_m$, the estimated variance is calculated as $s^{adj}_{m} = P(s_{m}| s_\Theta)$ where $s_{m}$ is the variance resulting from the linear model of PTM $m$ and $s_\Theta$ is the global variance across all PTMs. The statistical significance is determined using $\hat{\Delta}_m$ and $\widehat{SE(\hat{\Delta})}$ calculated using the adjusted variance $s^{adj}_{m}$.

@@

The proposed approach leverages the large amount of work done to create $MSstats$~\cite{Choi:2014}. $MSstats$ was designed for protein significance analysis and statistical relative quantification of proteins and peptides in global, targeted and data-independent proteomics. In terms of Figure \ref{fig:data-structure}, $MSstats$ was designed to estimate the difference in the unmodified protein in condition $1$ and condition $2$, denoted by $\mu_1$ and $\mu_2$ respectively. In other words it allows us to answer the new Null Hypothesis seen in Equation \ref{eq:msstats_null_hyp}.

@@

We propose a statistical method for detecting changes in PTMs that allows us to answer the Null Hypothesis stated in Equation \ref{eq:null_hyp}; there is no difference in log-abundance of the PTM site between conditions, adjusted with respect to unmodified protein abundance. It leverages the methods used in $MSstats$ to answer this hypothesis, while being applicable to experiments with complex designs. Relating to the example in Figure \ref{fig:data-structure}, the expected value of the log-abundance of the PTM and unmodified protein in Condition $1$ are denoted by $\hat{\mu}_{1}^{\ast}$ and $\hat{\mu}_{1}$ respectively. The feature and run level summarization for Condition $1$ is denoted as $\hat{y}_{1++}^{\ast}$ for the modified peptide and $\hat{y}_{1++}$ for the unmodified peptide.

To calculate $\hat{y}_{1++}^{\ast}$ and $\hat{y}_{1++}$, feature and run level summarization is carried out as in the sub-plot model of MSstats using Tukey's median polish. The summarized values for the PTM and unmodified protein are then used to fit two models based on the family of linear mixed-effects models. Finally, the adjusted difference, $\hat{\Delta}$, is given by the difference in log-abundance of the PTM site, subtracted by the difference in log-abundance of the underlying protein, which is equivalent to the log of the ratio of PTM abundance difference to protein abundance difference.

\begin{equation}
\hat{\Delta} = \left[ \frac{1}{J} \left( \hat{y}_{1++}^{\ast} - \hat{y}_{2++}^{\ast} \right) \right] - \left[ \frac{1}{J} \left( \hat{y}_{1++} - \hat{y}_{2++} \right) \right]
\label{eq:prop_fc}
\end{equation}

@@

\begin{equation}
\widehat{SE(\Delta)} = \left[ \left( \hat{\sigma}_{\gamma^{\ast}}^{2} + \hat{\sigma}_{\gamma}^{2} \right) \right]^{1/2}
\label{eq:prop_se}
\end{equation}

@@

After using the converters, the next step is peptide/protein summarization and missing value imputation. For the modified run, the package summarizes features which include the same modification together. Features with multiple modification sites are not included with single site features and are summarized separately. For the global profiling run all unmodified features from the same protein are summarized together, up to the protein level. Additionally, the summarization includes global median normalization and normalizes between MS runs. The package can optionally impute missing values using an AFT model. Missing value imputation should be performed with care when applied to PTM data. PTMs generally exhibit low feature counts and may be missing due to reasons other than low abundance. These issues can cause the missing value imputation to be unstable.
